Replicator-mutator and noise-perturbed replicator dynamics

### 2014-09-30 paper/03_simulations.tex
comparison with QRE

\subsubsection{Results}
Figure~\ref{fig:MeanMetrics} shows the means of the recorded metrics,
together with their estimated 95\% confidence intervals, for each
triple of independent variables. Most trends are quite clear, and not
unexpected. Entropy of sender and receiver strategies are increasing
in imprecision. Entropy of speaker strategies seems independent of
tolerance levels and the number of states, while there seems to be an
effect of both parameters on the entropy of the hearer
strategy. Speaker informativity and expected utility are decreasing
under growing imprecision. There seem to be mild effects of the number
of states in the game, but hardly any in terms of pragmatic
tolerance. Interestingly, Voronoiness of both sender and receiver
strategies appear to be non-decreasing under increasing impairment and
peaking at the maximal value. The only exception is the 6 states games
and the transition from $\impairment = 0$ to $\impairment =
0.05$. Expected utility of evolved languages mostly decreases in
increasing imprecision, with the sole exception of a rise for $\toler
= 0.2$ and $ns = 6$ at $\impairment = 0.1$. Still, normalized expected
utility of evolved strategy pairs under mild values of imprecision are
not necessarily much worse than their respective perfect precision
\rd-analogues.

@@

@@

  \caption{Proportion of converged and convex trials for each triple
    of independent variables.}
  \label{fig:CategoricalMeasures}
\end{figure}

Diffusion of strategies from confusion of states has another
interesting regularizing effect on the evolution of signaling. If we
look at the confidence intervals in Figure~\ref{fig:MeanMetrics}, we
see that there is very little variation in the recorded metrics for
evolved strategies, at least for higher values of impairment. On
closer inspection, it turns out that variability in low-impairment
conditions is not only due to non-convergence and/or associated
non-convexity. Figure~\ref{fig:MoreExample_strats} gives two more
examples of strategy pairs at stopping time. Both are obtained for the
same triple of parameters, both converged and are convex. However,
they are not equally efficient. In fact, the pair in
Figure~\ref{fig:example_stratsC} has a normalized expected utility of
$0.99$ while the one in Figure~\ref{fig:example_stratsD} only has
$0.89$.

@@

Interestingly, this type of variability in evolutionary outcomes is
also weeded out by mild increases in impairment. To see this, we
calculated the average distance between evolved sender strategies
within each group of trials that had identical independent
parameters. We determined the distance between sender strategies
$\Sstrat$ and $\Sstrat'$ as the average Hellinger distance between
probability distributions $\Sstrat(\state)$ and $\Sstrat'(\state)$
over messages at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
      \\
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $Sstrat$ with reversed message indices. The
average distances between sender strategies in this sense is plotted
in Figure~\ref{fig:AverageEUinGroups}. This clearly shows that with
increasing imprecision, the sender strategies evolving under \rdd were
exactly alike, modulo which message was used for which part of the
unit interval. In other words, perceptual imprecision clearly unifies
evolutionary outcomes, and in fact guarantees sender strategies that
are not only convex, but also maximally efficient in that they induce
a vague category split exactly in the middle of the unit interval.

@@

    \caption{Mean group measures as defined in the main text. For each
      group, i.e., triplet of independent variables, the plot shows
      (i) average normalized expected utility when playing against an
      arbitrary other language from the group, (ii) the mean distance
      between evolved strategies in the group.}
  \label{fig:AverageEUinGroups}
\end{figure}

>>>>

The strategy pair in Figure~\ref{fig:example_stratsA} has a normalized
expected utility of $0.992$, the pair in
Figure~\ref{fig:example_stratsB} has $0.901$. Generally, vagueness and
imprecision can be expected to decrease expected utility
\citep{Lipman2009:Why-is-Language}. The crucial question is whether
communicative success drops unacceptably fast with moderate levels of
vagueness and imprecision.

\subsubsection{Results}

@@

Figure~\ref{fig:MeanMetrics} shows the means of the recorded metrics,
together with their estimated 95\% confidence intervals, for each
triple of independent variables. Most trends are quite clear, and not
unexpected. Entropy of sender and receiver strategies are increasing
in imprecision. Entropy of speaker strategies seems independent of
tolerance levels and the number of states, while there seems to be an
effect of both parameters on the entropy of the hearer
strategy. Speaker informativity and expected utility are decreasing
under growing imprecision. There seem to be mild effects of the number
of states in the game, but hardly any in terms of pragmatic
tolerance. What was not expected \emph{a priori} was that Voronoiness
of both sender and receiver strategies appear to be non-decreasing
under increasing impairment and peaking at the maximal value. The only
exception is the 6 states games and the transition from $\impairment =
0$ to $\impairment = 0.05$. Expected utility of evolved languages
mostly decreases in increasing imprecision, with the sole exception of
a rise for $\toler = 0.2$ and $ns = 6$ at $\impairment = 0.1$. Still,
normalized expected utility of evolved strategy pairs under mild
values of imprecision are not necessarily much worse than their
respective perfect precision \rd-analogues. This is especially so for
lower numbers of states (which increase the chance of non-convergence
and non-convexity; see Figure~\ref{fig:CategoricalMeasures} and the
discussion below).

@@

  \caption{Proportion of converged and convex trials.}
  \label{fig:CategoricalMeasures}
\end{figure}

Diffusion also has another interesting regularizing effect on the
evolution of signaling. If we look at the confidence intervals in
Figure~\ref{fig:MeanMetrics}, we see that there is very little
variation in the recorded metrics for evolved strategies, at least for
higher values of impairment. On closer inspection, it turns out that
variability in low-impairment conditions is not only due to
non-convergence and/or associated
non-convexity. Figure~\ref{fig:MoreExample_strats} gives two more
examples of strategy pairs at stopping time. Both are obtained for the
same triple of parameters, both converged and are convex. However,
they are not equally efficient. In fact, the pair in
Figure~\ref{fig:example_stratsC} has a normalized expected utility of
$0.99$ while the one in Figure~\ref{fig:example_stratsD} only has
$0.89$.

@@

Interestingly, this type of variability in evolutionary outcomes is
also weeded out by mild increases in impairment. To see this, we
calculated the average distance between evolved sender strategies
within each group of trials that had identical independent
parameter values. We determined the distance between sender strategies
$\Sstrat$ and $\Sstrat'$ as the average Hellinger distance between
probability distributions $\Sstrat(\state)$ and $\Sstrat'(\state)$
over messages at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. The
average distances between sender strategies in this sense is plotted
in Figure~\ref{fig:AverageEUinGroups}. This clearly shows that with
increasing imprecision, the sender strategies evolving under \rdd were
exactly alike, modulo which message was used for which part of the
unit interval. In other words, perceptual imprecision clearly unifies
evolutionary outcomes, and in fact guarantees sender strategies that
are not only convex, but also maximally efficient in that they induce
a vague category split exactly in the middle of the unit interval.

@@

    \caption{Mean group measures as defined in the main text. For each
      group, i.e., triplet of independent variables, the plot shows
      (i) average normalized expected utility when playing against an
      arbitrary other language from the group, (ii) the mean distance
      between evolved sender strategies in the group.}
  \label{fig:AverageEUinGroups}
\end{figure}

### 2014-09-30 paper/04_discussion.tex
simulation results write up
comparison with QRE

>>>>

\subsection{Relation with previous accounts}

\paragraph{Quantal response equilibria.}
\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in
which information processing limitations of signaling agents could
lead to vague strategies. The model that is most clearly related to
the present approach uses the notion of a quantal response, also known
as a logit response or a soft-max function
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,McKelveyPalfrey1995:Quantal-Respons,McKelveyPalfrey1998:Quantal-Respons,GoereeHolt2008:Quantal-Respons}. A
quantal response function is a paramterized generalization of the
classic best response function. For example, if $U \mycolon \Acts
\rightarrow \mathds{R}$ is the measure of expected utility over
choices $\Acts$ of an agent, then a best response function would have
the agent choose $\act$ only if $U(\act) = \max_{\act' \in \Acts}
U(a)$. A quantal response function rather assumes that agents would
choose $\act$ with a probability proportional to $\expo(\lambda \cdot
U(\act))$, where $\lambda$ is a rationality parameter. If $\lambda
\rightarrow \infty$ we retrieve the behavior of the best response
choice function, but if it is positive but finite, any choice $\act$
will receive a positive probability of being chosen, but acts with
higher expected utility will be more
likely. \citet{FrankeJager2010:Vagueness-Signa} show that quantal
response equilibria of sim-max games, i.e., pairs of sender and
receiver strategies such that the sender strategy is the quantal
response to the expected utilities under the receiver strategy and
vice versa, show the desired marks of vague
signaling. Figure~\ref{fig:exampleQRE_stratsA} shows an example of a
quantal response equilibrium for a sim-max game, as used in our set-up
with $\toler = 0.5$. Sender and receiver strategies look very much
like what evolves under \rdd with modest values of perceptual
imprecision.

There is a major conceptual difference in the source of vagueness
between the quantal response approach and our favored approach in
terms of perceptual noise. When we choose a lower level of pragmatic
tolerance, like we used in our simulations, the predictions of the
quantal response approach look
different. Figure~\ref{fig:exampleQRE_stratsB} shows such an example
that clearly shows that, unlike the evolving strategies under \rdd,
sender strategies have vague boundaries also towards the end of the
unit intervals. This is because quantal responses equalize message use
far away from the ``prototypical'' interpretation. That this is
conceptually odd shows even more clearly in a case where the state
space is intuitively unbounded, as for instance for the property
``tall''. If the usual interpretation of a ``tall man'' peaks at
around, say, 195cm then when meeting a giant of $n$ meters speaker
would, according to the quantal response approach, be ever more
inclined to describe the giant as ``short''. This is because, as the
distance from the prototype increases for larger $n$, the expected
utilities of saying ``short'' or ``tall'' will both converge to zero,
and the difference between the two will also converge to zero. Whence
that the quantal response approach would predict that speakers would
grow indifferent between choice of antonyms as $n$ grows, which seems
weird. (Notice that this argument hinges on the choice of utility
function. Still, to the extent that the chosen lower bounded utility
functions are reasonable ---and we think they are very reasonable---,
this is a conceptual oddity.)

The \rdd approach does not have this peculiarity. \dots

\begin{figure}
  \centering

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/exampleStratQRE_tolerance05.pdf}
    \caption{$\ns = 90$, $\lambda = 15$, $\toler = 0.5$}
    \label{fig:exampleQRE_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/exampleStratQRE_tolerance005.pdf}
    \caption{$\ns = 90$, $\lambda = 15$, $\toler = 0.05$}
    \label{fig:exampleQRE_stratsB}
  \end{subfigure}

  \caption{Examples of vague quantal response equilibria.}
  \label{fig:exampleQREs}
\end{figure}

### 2014-09-30 snippets/data_exploration/data_exploration_tmp.Rmd
comparison with QRE

pd <- position_dodge(.01)
sp = ggplot(dsub, aes(x=Impairment, y=value, colour=factor(Number.of.states))) + 
     geom_point(position = pd) +
     geom_errorbar(aes(ymin=value-ci, ymax=value+ci), width=0, position=pd) +
     geom_line(position=pd)

@@

pdf("~/Desktop/data/svn/vagueness-games/paper/plots/MeanMetrics.pdf", width=8, height = 8)
show(sp)
dev.off()

@@

sp = ggplot(dms, aes(x=factor(Impairment), y = value)) 
sp = sp + geom_bar(aes(fill = variable), color = "black", stat="identity", position= position_dodge(0.4), width = 0.3)
sp = sp + facet_grid(Number.of.states ~ Tolerance, scales="free")

@@

pdf("~/Desktop/data/svn/vagueness-games/paper/plots/CategoricalMeasures.pdf", width=8, height = 6)
show(sp)
dev.off()

>>>>

dms = summarySE(dm, groupvars = c("Number.of.states", "Impairment", "Tolerance", "variable"), measurevar="value")
dsub = dms
dsub$Number.of.states = factor(dsub$Number.of.states)

pd <- position_dodge(.01)
sp = ggplot(dsub, aes(x=Impairment, y=value, colour=Number.of.states)) + 
     geom_point(position = pd) +
     geom_errorbar(aes(ymin=value-ci, ymax=value+ci), width=0, position=pd) +
     geom_line(position=pd) + xlab("imprecision")

@@

pdf("~/Desktop/data/svn/vagueness-games/paper/plots/MeanMetrics.pdf", width=8, height = 6)
show(sp)
dev.off()

@@

sp = ggplot(dms, aes(x=factor(Impairment), y = value)) 
sp = sp + geom_bar(aes(fill = variable), color = "black", stat="identity", position= position_dodge(0.4), width = 0.35)
sp = sp + facet_grid(Number.of.states ~ Tolerance, scales="free") + ylab("proportion") + xlab("imprecision")

@@

pdf("~/Desktop/data/svn/vagueness-games/paper/plots/CategoricalMeasures.pdf", width=8, height = 5)
show(sp)
dev.off()
